---
title: "Linear model with multiple predictors <br> `r emo::ji('woman_juggling')`"
author: "Dr. Çetinkaya-Rundel"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(broom)
library(here)
library(plotly)
library(widgetframe)
```

```{r load-pp, include=FALSE}
pp <- read_csv(here::here("csv/paris-paintings.csv"), na = c("n/a", "", "NA"))
```


## Announcements

- Student hours later today:
  - Wed, 14:30 - 15:30 @ [bit.ly/ids-zoom-week-08](http://bit.ly/ids-zoom-week-08) (Online!)
- ...

<br>

---

class: center, middle

# The linear model with multiple predictors

---

## Multiple predictors

- Response variable: log(price)
- Explanatory variables: Width and height

```{r model-price-width-height}
m_wi_hgt <- lm(log(price) ~ Width_in + Height_in, data = pp)
tidy(m_wi_hgt)
```

---

##  Linear model with multiple predictors

```{r model-price-width-height-tidy, echo=FALSE}
tidy(m_wi_hgt)
```

$$\widehat{log(price)} = 4.77 + 0.0269 width - 0.0133 height$$

---

## Visualizing models with multiple predictors

```{r plotly, echo=FALSE, warning=FALSE, cache=TRUE}
p <- plot_ly(pp, x = ~Width_in, y = ~Height_in, z = ~log(price),
        marker = list(size = 3,
                       color = "lightgray",
                       alpha = 0.5,
                       line = list(color = "gray",
                                   width = 2))) %>%
  add_markers() %>%
  plotly::layout(scene = list(xaxis = list(title = "Width (in)"),
                      yaxis = list(title = "Height (in)"),
                      zaxis = list(title = "Log(price)"))) %>%
  config(displayModeBar = FALSE)
frameWidget(p, width = "100%", height = "100%")
```

---

## Exploration 1

### Price, surface area, and living artist

- Explore the relationship between price of paintings and surface area, conditioned 
on whether or not the artist is still living
- First visualize and explore, then model

---

## Typical surface area

.question[
What is the typical surface area for paintings?
]

```{r viz-surf-artistliving, echo=FALSE,warning=FALSE, fig.width=5, fig.height=2}
ggplot(data = pp, 
       mapping = aes(y = log(price), x = Surface, color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  labs(color = "Living artist")
```

--

Less than 1000 square inches (~ 80cm x 80cm). There are very few paintings that have surface area above 5000 square inches.

---

## Narrowing the scope

For simplicity let's focus on the paintings with `Surface < 5000`:

```{r surf-lt-5000}
pp_Surf_lt_5000 <- pp %>%
  filter(Surface < 5000)
```

```{r viz-surf-lt-5000-artistliving, echo=FALSE, warning=FALSE, fig.width=5, fig.height=1.8}
ggplot(data = pp_Surf_lt_5000, 
       mapping = aes(y = log(price), x = Surface, color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  labs(color = "Living artist")
```

---

## Two ways to model

- **Main effects:** Assuming relationship between surface and logged price 
**does not vary** by whether or not the artist is living.
- **Interaction effects:** Assuming relationship between surface and logged 
price **varies** by whether or not the artist is living.

.pull-left[
```{r viz-main-effects, fig.height=3.8, echo=FALSE}
m_main <- lm(log(price) ~ Surface + factor(artistliving), data = pp_Surf_lt_5000)
m_main_aug <- augment(m_main)
ggplot(data = m_main_aug, 
       mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted), size = 1.2) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist", 
       title = "Main effects") +
  theme_minimal()
```
]
.pull-right[
```{r viz-interaction-effects, fig.height=3.8, echo=FALSE}
ggplot(data = pp_Surf_lt_5000,
       mapping = aes(y = log(price), x = Surface, 
                     color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(y = "Log(price)", color = "Living artist", title = "Interaction effects") +
  theme_minimal()
```
]

---

## Fit model with main effects

- Response variable: log(price)
- Explanatory variables: Surface area and artist living (0/1 variable)

.midi[
```{r model-main-effects}
m_main <- lm(log(price) ~ Surface + factor(artistliving), 
             data = pp_Surf_lt_5000)
tidy(m_main)
```
]


--
- Linear model:
$$ \widehat{log(price)} = 4.88 + 0.000265~surface + 0.137~artistliving $$

---

## Solving the model

- Non-living artist: Plug in 0 for `artistliving`
$\widehat{log(price)} = 4.88 + 0.000265~surface + 0.137 \times 0$  
$= 4.88 + 0.000265~surface$

--
- Living artist: Plug in 1 for `artistliving`
$\widehat{log(price)} = 4.88 + 0.000265~surface + 0.137 \times 1$  
$= 5.017 + 0.000265~surface$

---

## Visualizing main effects

```{r fig.height=1.75, echo = FALSE}
ggplot(data = m_main_aug, 
       mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist") +
  theme_minimal()
```

- **Same slope:** Rate of change in price as the surface area increases does 
not vary between paintings by living and non-living artists.
- **Different intercept:** Paintings by living artists are consistently more 
expensive than paintings by non-living artists.

---

## Interpreting main effects

.midi[
```{r exp-coefs}
tidy(m_main) %>% 
  mutate(exp_estimate = exp(estimate)) %>%
  select(term, estimate, exp_estimate)
```
]

- All else held constant, for each additional square inch in painting's surface area, the price of the painting is predicted, on average, to be higher by a factor of 1.
- All else held constant, paintings by a living artist are predicted, on average, to be higher by a factor of 1.15 compared to paintings by an artist who is no longer alive.
- Paintings that are by an artist who is not alive and that have a surface area of 0 square inches are predicted, on average, to be 132 livres.

---

## What went wrong?

.question[
Why is our linear regression model different from what we got from `geom_smooth(method = "lm")`?
]

.pull-left[
```{r echo=FALSE, fig.height=4}
ggplot(pp_Surf_lt_5000, aes(x = Surface, y = log(price), color = factor(artistliving))) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```
]
.pull-right[
```{r viz-main-effects3, echo=FALSE, fig.height=4}
m_pr <- lm(log(price) ~ Surface + factor(artistliving), data = pp_Surf_lt_5000)
m_pr_aug <- augment(m_pr)
ggplot(data = m_pr_aug, mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```
]

---

## What went wrong? (cont.)

- The way we specified our model only lets `artistliving` affect the intercept.
- Model implicitly assumes that paintings with living and deceased artists have the *same slope* and only allows for *different intercepts*.  
- What seems more appropriate in this case? 
    + Same slope and same intercept for both colors
    + Same slope and different intercept for both colors
    + Different slope and different intercept for both colors?

---

## Interacting explanatory variables

- Including an interaction effect in the model allows for different slopes, i.e. 
nonparallel lines.
- This implies that the regression coefficient for an explanatory variable would 
change as another explanatory variable changes.
- This can be accomplished by adding an interaction variable: the product of two 
explanatory variables.

---

## Interaction: surface * artist living

```{r fig.height=1.75, echo = FALSE}
m_int <- lm(log(price) ~ Surface * factor(artistliving), data = pp_Surf_lt_5000)
m_int_aug <- augment(m_int)
ggplot(data = m_int_aug, 
       mapping = aes(y = log.price., x = Surface, color = factor.artistliving.)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = .fitted)) +
  labs(x = "Surface", y = "Log(price)", color = "Living artist") +
  theme_minimal()
```

---

.small[
```{r viz-interaction-effects1, fig.height=1.75}
ggplot(data = pp_Surf_lt_5000,
       mapping = aes(y = log(price), x = Surface, 
                     color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(x = "Surface", y = "Log(price)", color = "Living artist")
```
]

---

## Fit model with interaction effects

- Response variable: log(price)
- Explanatory variables: Surface area, artist living (0/1 variable), and 
their interaction

.midi[
```{r model-interaction-effects}
m_int <- lm(log(price) ~ Surface + factor(artistliving) + 
              Surface * factor(artistliving), 
            data = pp_Surf_lt_5000)
tidy(m_int)
```
]

- Linear model:
$$ \widehat{log(price)} = 4.91 + 0.00021~surface - 0.126~artistliving $$
$$+ ~ 0.00048~surface \times artistliving $$

---

## Interpretation of interaction effects

- Rate of change in price as the surface area of the painting increases does 
vary between paintings by living and non-living artists (different slopes), 
- Some paintings by living artists are more expensive than paintings by
non-living artists, and some are not (different intercept).

.small[
.pull-left[
- Non-living artist: 
$\widehat{log(price)} = 4.91 + 0.00021~surface$
$- 0.126 \times 0 + 0.00048~surface \times 0$
$= 4.91 + 0.00021~surface$
- Living artist: 
$\widehat{log(price)} = 4.91 + 0.00021~surface$
$- 0.126 \times 1 + 0.00048~surface \times 1$
$= 4.91 + 0.00021~surface$
$- 0.126 + 0.00048~surface$
$= 4.784 + 0.00069~surface$
]
.pull-right[
```{r viz-interaction-effects2, fig.height=4, echo = FALSE}
ggplot(data = pp_Surf_lt_5000,
       aes(y = log(price), x = Surface, color = factor(artistliving))) +
  geom_point(alpha = 0.3) +
  stat_smooth(method = "lm", fullrange = TRUE)
```
]
]

---

## Third order interactions

- Can you? Yes
- Should you? Probably not if you want to interpret these interactions in context
of the data.
