---

class: center, middle

# Exploring linearity

---

## Data: Paris Paintings

```{r echo=FALSE}
ggplot(data = pp, aes(x = price)) +
  geom_histogram(binwidth = 1000) +
  labs(title = "Prices of paintings")
```

---

## Price vs. width

.question[
Describe the relationship between price and width of painting.
]

```{r echo=FALSE, fig.height=2.75}
ggplot(data = pp, aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(x = "Width (in)", y = "Price (livres)")
```

---

## Let's focus on paintings with `Width_in < 100`

```{r}
pp_wt_lt_100 <- pp %>% 
  filter(Width_in < 100)
```

---

## Price vs. width

.question[
Which plot shows a more linear relationship?
]

.small[
  
.pull-left[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(title = "Price vs. width", subtitle = "For width < 100 in",
       x = "Width (in)", y = "Price (livres)")
```
]

.pull-right[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  labs(title = "Log(price) vs. width", subtitle = "For width < 100 in",
       x = "Width (in)", y = "Log(price) (log livres)")
```
]

]

---

## Price vs. width, residuals

.question[
Which plot shows a residuals that are uncorrelated with predicted values from the model?
]
  
.pull-left[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
m_wi_pr <- lm(price ~ Width_in, data = pp_wt_lt_100)
m_wi_pr_tidy <- augment(m_wi_pr)
ggplot(data = m_wi_pr_tidy, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Price vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted price (livres)", 
    y = "Residuals"
    )
```
]
.pull-right[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
m_log_wi_pr <- lm(log(price) ~ Width_in, data = pp_wt_lt_100)
m_log_wi_pr_tidy <- augment(m_log_wi_pr)
ggplot(data = m_log_wi_pr_tidy, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Log(Price) vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted log(price) (log livres)", 
    y = "Residuals"
    )
```
]

--

<br>
.question[
What is the unit of residuals?
]

---

## Transforming the data

- We saw that `price` has a right-skewed distribution, and the relationship between price and width of painting is non-linear.

--
- In these situations a transformation applied to the response variable may be useful.

--
- In order to decide which transformation to use, we should examine the distribution of the response variable.

--
- The extremely right skewed distribution suggests that a log transformation may 
be useful.
    - log = natural log, $ln$
    - Default base of the `log` function in R is the natural log: <br>
    `log(x, base = exp(1))`
    
---

## Logged price vs. width

.question[
How do we interpret the slope of this model?
]

```{r echo=FALSE}
ggplot(data = pp_wt_lt_100, mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "#A7D5E8", se = FALSE) +
  labs(x = "Width (in)", y = "Log(price) (log livres)")
```

---

## Interpreting models with log transformation

```{r}
m_lprice_wt <- lm(log(price) ~ Width_in, data = pp_wt_lt_100)
tidy(m_lprice_wt)
```

---

## Linear model with log transformation

$$ \widehat{log(price)} = 4.67 + 0.02 Width $$

--
- For each additional inch the painting is wider, the log price of the
painting is expected to be higher, on average, by 0.02 livres.

--
- which is not a very useful statement...

---

## Working with logs

- Subtraction and logs: $log(a) − log(b) = log(a / b)$

--
- Natural logarithm: $e^{log(x)} = x$

--
- We can use these identities to "undo" the log transformation

---

## Interpreting models with log transformation

The slope coefficient for the log transformed model is 0.02, meaning the log 
price difference between paintings whose widths are one inch apart is predicted 
to be 0.02 log livres.

--

$$ log(\text{price for width x+1}) - log(\text{price for width x}) = 0.02 $$

--

$$ log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right) = 0.02 $$

--

$$ e^{log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right)} = e^{0.02} $$

--

$$ \frac{\text{price for width x+1}}{\text{price for width x}} \approx 1.02 $$

--

For each additional inch the painting is wider, the price of the
painting is expected to be higher, on average, by a factor of 1.02.

---

## Shortcuts in R

```{r}
m_lprice_wt %>%
  tidy() %>%
  select(term, estimate)
```

```{r}
m_lprice_wt %>%
  tidy() %>%
  select(term, estimate)
```

---

## Recap

- Non-constant variance is one of the most common model violations, however it 
is usually fixable by transforming the response (y) variable.

--
- The most common transformation when the response variable is right skewed is 
the log transform: $log(y)$, especially useful when the response variable is 
(extremely) right skewed.

--
- This transformation is also useful for variance stabilization.

--
- When using a log transformation on the response variable the interpretation of 
the slope changes: *"For each unit increase in x, y is expected on average to be higher/lower <br> by a factor of $e^{b_1}$."*

--
- Another useful transformation is the square root: $\sqrt{y}$, especially 
useful when the response variable is counts.

---

## Transform, or learn more?

- Data transformations may also be useful when the relationship is non-linear
- However in those cases a polynomial regression may be more appropriate
  + This is beyond the scope of this course, but you’re welcomed to try it for your final project, and I’d be happy to provide further guidance

---

## Aside: when $y = 0$

In some cases the value of the response variable might be 0, and

```{r}
log(0)
```

--

The trick is to add a very small number to the value of the response variable for these cases so that the `log` function can still be applied:

```{r}
log(0 + 0.00001)
```
